---
title: "Class 8 Mini-Project"
author: "Merna Behnam (PID: A17675170)"
format: pdf
toc: true
---
## Background

In today's class we will apply the methods and techniques clustering and PCA to help make sense of a real world breast cancer **FNA (fine needle aspirations)** biopsy data set.

## Data Import

We start by importing our data. It is a CSV file, so we will use the `read.csv()` function. First, download the file containing the data, and place it in the project for this class on your computer. 

```{r}
#read.csv("WisconsinCancer.csv")
```

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names = 1)
```

have a wee peek at the first few rows:
```{r}
head(wisc.df, 4)
```

Omit the first column `diagnosis` because I don't want to use this for my machine learning models. We will use it later on to comapre our results to the expert diagnosis:
```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]

# Create diagnosis vector for later 
diagnosis <- wisc.df$diagnosis
```

Make sure to remove the first `diagnosis` column - I don't want to use this for my machine learning models. We will see it later to compare our results to the expert diagnosis. 

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Exploratory data analysis

> Q1. How many observations are in this dataset?

```{r}
nrow((wisc.data))
```

There are a total of 569 observations in this dataset.

> Q2. How many of the observations have a malignant diagnosis?

```{r}
#Method 1
sum(wisc.df$diagnosis == "M")
```

```{r}
#Method 2
table(wisc.df$diagnosis)
```

There are 212 malignant diagnoses. 

> Q3. How many variables/features in the data are suffixed with `_mean`?

```{r}
#colnames(wisc.data)
length(grep("_mean", colnames(wisc.data)))
```

There are 10 variables in the dataset with the suffic `_mean`.

## Principal Component Analysis

The main function here is `prcomp()` and we want to make sure we set the optional argument `scale=TRUE`:
```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

44.27% of the original variance is captured by PC1.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Three PCs are required to describe at least 70% of the original variance in the data.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

Seven PCs are required to describe at least 90% of the original variance in the data.

### Interpreting PCA Results

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

The plot is incredibly difficult to read, and I have no idea what's going on. This is not an optimal way to plot and clearly interpret our data.

Our main PCA "score plot" or "PC plot" of results:
```{r}
library(ggplot2)
```
```{r}
ggplot(wisc.pr$x) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

This plot is a lot more effective in communicating the difference between malignant and benign diagnoses.

scree-plot:
```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

Collectively these two plot ("score plot" and "loading plot") tell us that if cells nucleus are deeply indented ("concave"), irregular and non circular ("compactness"), and have large "perimeter" values they tend to be malignant...

> Q9. For the first principal component, what is the component of the loading vector (i.e. `wisc.pr$rotation[,1]`) for the feature `concave.points_mean`? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

## Heirarchal Clustering

First scale the data (with the `scale()` function), then calculate a distance matrix (with the `dist()` function), then cluster with the `hclust()` function and plot:

```{r}
wisc.hclust <- hclust( dist( scale( wisc.data)))
```

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=20, col="red")
```

You can also use the `cutree()` function with a argument `k=4` rather than `h=height`.

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters)
```

## Combining Methods

> Q11. OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?

Here we will take our PCA results and use those as input for clustering. In other words our `wisc.pr$x` scores that we plotted above (the main output from PCA - how the data lie on our new principal component axis/variables) and use a subset of these PCs that capture the most variance as input for `hclust()`.

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

I *much* prefer the PCA plot over the dendrogram. I genuinely can not understand the dendrogram and it's too crowded.

Cut the dendrogram/tree into two main groups/clusters:

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

I want to know how the clustering in `grps` with values of 1 or 2 correspond the expert `diagnosis`.

> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
table(grps, diagnosis)
```

My clustering **group 1** are mostly "M" diagnosis (179) and y clustering **group 2** are mostly "B" diagnosis.

24 False Positives
179 True Positives
333 True Negatives
33 False Negatives

Sensitivity TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity TN/(TN+FP)
```{r}
333/(333+24)
```

> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (`wisc.hclust.clusters` and `wisc.pr.hclust.clusters`) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
```

## Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

We would most likely want to follow up with the patients that had a malignant biopsy (black points on the plot). The severe cases are seen in the extremeties of the plot, and should also be considered for priority. 
