---
title: "Class 12 Homework: Pt.2 (Population Analysis) [Q13 & Q14 Boxplot]"
author: "Merna Behnam (PID: A17675170)"
format: pdf
---

Section 4: Population Scale Analysis [HOMEWORK]

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. 

How many Samples do we have?

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(dplyr)

expr %>%
  group_by(geno) %>%
  summarise(median_expression = median(exp))
```

The total sample size is 462. The sample size of G]genotype A|A is 108, A|G with 233, and G|G with 121. The median expression level for A|A is 31.25, A|G at 25.06, and G|G at 20.07.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

```{r}
library(ggplot2)
```

Let's make a boxplot:

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
```

Based on the boxplot, its clear that having a G|G in this location has significantly reduced expression compared to A|A in this gene. This suggests that the SNP affects ORMDL3 expression, since different genotypes are associated with distinct expression levels, consistent with a regulatory effect.






















