---
title: "Class 19: Cancer Mutation Project"
author: "Merna Behnam (PID: A17675170)"
format: html
---

I downloaded my sequence from the class website and moved it to my working directory:

From a **BLASTp** search against human refseq I see that it is

- Official Symbol: TP53
- Official Full Name: cellular tumor antigen p53 isoform a

```{r}
library (bio3d)

a <- read.fasta("A17675170_mutant_seq.fa")
a
```



We could score residue conservation then find the non 1.0 scoring positins. These will be the mutation positions:

```{r}
conserv(a)
```

```{r}
mutation.sites <- which(conserv(a)<1)
```

```{r}
paste( a$ali[1, mutation.sites],
       mutation.sites,
       a$ali[2, mutation.sites],sep=""
)
```

Domains from hmmer search

![] (hmmer_domains.png)

PFAM: 
Description:
Coordinates:


only healthy wt:
```{r}
wt <- a$ali["wt_healthy", ]
wt_seq <- paste(wt, collapse = "")
wt_seq
```

```{r}
hits <- blast.pdb(wt)

hits$hit.tbl[1:10, ]
```

```{r}
rownames(a$ali)
```

only tumor:
```{r}
mt <- a$ali["mutant_tumor", ]
mt_seq <- paste(mt, collapse = "")
mt_seq
```





















