---
title: "Class 17: Analyzing sequencing data in the cloud"
author: "Merna Behnam (PID: A17675170)"
format: pdf
---

## Background
In bioinformatics we often need to analyze data-sets that are too large (or would take too long) to analyze on our local lab computers. The goal of this hands-on session is to show you how to obtain and work with large biomolecular sequencing data-sets using cloud computing resources. In particular, we will query, download, decompress and analyze data from NCBI’s main Sequence Read Archive (SRA) the world’s largest publicly available repository of high throughput sequencing data.

## The Sequence Read Archive

> Q. What shell command can you use to view the top few files of your FASTQ file?

`head SRR600956.fastq`

> Q. What length are these sequence reads?

38

> Q. Can you use the grep command to determine how many total reads are in this file?

Yes, using `grep -c "@SRR600956" SRR600956.fastq`.

> Q. Does your number of reads from grep match the name of the last read in the file? If not why not?

No, because some reads may be missing or filtered, so the numbering does not always match the exact count.


## Working with RNA-Seq data

> Q. How would you check that these files actually look like FASTQ?

`head SRR2156848_1.fastq`

> Q. How could you check the number of sequences in each file?

`grep -c "^@" SRR2156848_1.fastq`

> Q. Q. Check your answer with the bottom of the file using the tail command and also check the matching mate pair FASTQ file. Do these numbers match? If so why or why not?

Yes! They match because paired-end reads produce one read in each file for every sequencing spot, so both FASTQ files have the same number of reads.

> Q. Q. Check you have pairs of FASTQ files for all four datasets and that they have the same number of counts in each pair?

`ls *.fastq`
`grep -c "^@" *.fastq`


## Transcript quantification via pseudoalignment

> Q. Can you run kallisto to print out its citation information?

`kallisto cite`


> Q. Now run the quantification calculation for the remaining samples:

`kallisto quant -i hg19.ensembl -o SRR2156849_quant SRR2156849_1.fastq SRR2156849_2.fastq`
`kallisto quant -i hg19.ensembl -o SRR2156850_quant SRR2156850_1.fastq SRR2156850_2.fastq`
`kallisto quant -i hg19.ensembl -o SRR2156851_quant SRR2156851_1.fastq SRR2156851_2.fastq`


## Downstream analysis

There is an R function called `tximport()` in the tximport package, which enables straightforward import of Kallisto results

```{r}
library(tximport)

# setup the folder and file-names to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

```{r}
colSums(txi.kallisto$counts)
```

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```



## Principal Component Analysis

We can now apply any exploratory analysis technique to this counts matrix. As an example, we will perform a PCA of the transcriptomic profiles of these samples.

Now we compute the principal components, centering and scaling each transcript’s measured levels so that each feature contributes equally to the PCA:

```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

> Q. Use ggplot to make a similar figure of PC1 vs PC2 and a separate figure PC1 vs PC3 and PC2 vs PC3.

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```


## OPTIONAL: Differential-expression analysis

```{r}
library(DESeq2)
```

```{r}
sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)
```

```{r}
dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable, 
                                ~condition)
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
head(res)
```
